---
title: "chapter-clustering-validation"
author: "Ar"
date: '2022-07-19'
output: pdf_document
---

# Need for validation of clusters 
Clustering algorithms are designed such that they come out with a given number of clusters even if the underlying data is devoid of any such clusters.
We will see a criterion to assess the credibility of the clusters produced by any clustering algorithm.

## Within-groups sum of squared distances (WSS): 

$$
WSS_k =  \sum_{l=1}^k{\sum_{x_i \in {C_l}}{d^2(x_i, \overline{x_l})}}
$$

where, $k$ is the number of clusters and within the l-th cluster $C_l$, $x_l$ is the centre of mass.
We are interested in finding the *elbow* where there is a sudden drop in WSS_k as k is increased. 

```{r}
library("dplyr")
simdat = lapply(c(0,8), function(mx){

lapply(c(0,8), function(my){
tibble( x = rnorm(100, mean=mx, sd=2),
        y  = rnorm(100, mean=my, sd=2),
        class=paste(mx, my, sep=":"))

}) %>% bind_rows

}) %>% bind_rows
simdat
```

```{r}
library("tidyverse")
simdatxy = simdat[, c("x", "y")]
ggplot(simdat, aes(x=x, y=y, col=class))+geom_point()+coord_fixed()
```

```{r}
wss = tibble(k = 1:8, value = NA_real_)
wss$value[1] = sum(scale(simdatxy, scale=FALSE)^2)
for(i in 2:nrow(wss)){
km = kmeans(simdatxy, centers = wss$k[i])
wss

}
```

```{r}
?tibble()
```